---
title: "Calendar and holiday regressors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calendar and holiday regressors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  eval = FALSE
)
```

Calendar effects are part of the model specification. In `seasight`, they are
therefore treated as explicit candidate choices rather than hidden package data.
This keeps the package focused on model search, diagnostics and reporting while
leaving production calendars under the control of the user or institution.

The practical rule is:

- use X-13 or `seasonal` built-ins for standard effects where possible;
- use `build_user_xreg()` for institution-supplied moving-holiday dates;
- pass the resulting named regressors through `td_candidates`;
- inspect `with_td`, `td_name` and the report's "TD regressor" column.

`seasight` intentionally does not ship broad holiday datasets such as
`holiday` or `chcal`. National holidays, bridge days, school calendars,
religious holidays and trading calendars are source- and sector-specific. A
generic bundled calendar can easily be wrong for a production series.

## Built-in Easter handling

For Easter, the simplest path is usually to let X-13 handle the effect through
the model specification. `seasight` exposes this through `include_easter`.

```{r easter-built-in}
library(seasight)
library(seasonal)

y <- AirPassengers

res_easter <- auto_seasonal_analysis(
  y = y,
  include_easter = "auto",
  engine = "auto"
)

res_easter$table[1, c("model_label", "with_easter", "with_td", "td_name", "score_100")]
```

Use `include_easter = "always"` only when the effect is part of the production
policy for that series. Use `include_easter = "off"` when Easter is known to be
irrelevant or is handled elsewhere.

## Precomputed trading-day regressors

For working-day or trading-day effects, many institutions already maintain
calendar regressors. Pass those directly as a named list. The names are carried
through into the candidate table and HTML report.

```{r precomputed-td}
dates <- seq(as.Date("1949-01-01"), by = "month", length.out = length(y))

days_in_month <- function(date) {
  first <- as.Date(format(date, "%Y-%m-01"))
  next_month <- seq(first, by = "month", length.out = 2)[2]
  as.integer(next_month - first)
}

length_of_month <- vapply(dates, days_in_month, integer(1))
td_len <- ts(length_of_month - mean(length_of_month), start = start(y), frequency = frequency(y))

res_td <- auto_seasonal_analysis(
  y = y,
  td_candidates = list(length_of_month = td_len),
  td_usertype = "td",
  include_easter = "auto",
  engine = "auto"
)

res_td$table[1, c("model_label", "with_td", "td_name", "td_p", "score_100")]
```

The example above is deliberately simple. In production, the regressor should
come from the relevant business, statistical or national-accounts calendar.

## Moving holidays with `build_user_xreg()`

For holidays with known dates, such as Diwali, Lunar New Year, Ramadan-related
effects or local events, use `build_user_xreg()`. Internally this uses
`seasonal::genhol()` and then aligns the resulting regressor to the target
series.

```{r moving-holiday}
diwali_dates <- as.Date(c(
  "2018-11-07", "2019-10-27", "2020-11-14", "2021-11-04",
  "2022-10-24", "2023-11-12", "2024-11-01"
))

cny_dates <- as.Date(c(
  "2018-02-16", "2019-02-05", "2020-01-25", "2021-02-12",
  "2022-02-01", "2023-01-22", "2024-02-10"
))

holiday_xreg <- build_user_xreg(
  y = y,
  holidays = list(
    diwali = list(dates = diwali_dates, start = -7, end = 0),
    lunar_new_year = list(dates = cny_dates, start = -3, end = 3)
  ),
  td_usertype = "holiday"
)

names(holiday_xreg)

res_holiday <- auto_seasonal_analysis(
  y = y,
  td_candidates = holiday_xreg,
  include_easter = "auto",
  engine = "auto"
)

res_holiday$table[1, c("model_label", "with_td", "td_name", "with_easter", "td_p", "score_100")]
```

For SEATS candidates with user regressors, X-13 needs future regressor values
over its forecast horizon. `seasight` pads pulse-type holiday regressors with
zeros for this purpose. For non-pulse regressors, such as trading-day counts,
provide future calendar values yourself where possible.

## Inspecting the final decision

The candidate table and the HTML report show which calendar option was selected.
The final "Top candidates" table includes a "TD regressor" column with either the
candidate name, such as `diwali`, or `none`.

```{r report-with-holiday}
sa_report_html(
  y = y,
  td_candidates = holiday_xreg,
  include_easter = "auto",
  title = "AirPassengers calendar-regressor example",
  outfile = tempfile(fileext = ".html")
)
```

This makes the calendar decision auditable: a reviewer can see whether no
calendar regressor was used, whether Easter was included, and which user-supplied
holiday candidate won.

